Agile Formality: A Mole of Software Verification Practices

### 2010-07-02 papers/ModelBasedTestCaseGenerationDataStreamProtocol/paper.tex
rewrote remaining part of the introduction

@@

This software driver has been developed using our process while
statisfying seemingly contradictory requirements: (1) the protocol
implementation must be thoroughly verified; (2) parts of the protocol
must be formally specified; (3) a board simulator must be designed,
implemented, and verified to match the specifications of, and actual
behavior of, the physical circuit board; (4) if board components
change during development, the associated formal specifications and
the simulator whose behavior reflects such changes must be updated
accordingly; (5) strict software delivery deadlines.  Consequently,
these process requirements constitute an environment that benefits
from both an agile and a formal approaches to software development.

@@

@@

\subsection{A Real and Complex Case Study: \\ Rapid Development in Small-scale Hardware Software Co-design}
\label{subsec:a_real_and_complex_case_study}

% SenseTile introduction
The \textbf{UCD CASL SenseTile System} is a large-scale, general-purpose sensor system developed at the University College Dublin in Dublin, Ireland.
Our facility provide a scalable and extensible infrastructure for performing in-depth investigation of both the specific and general issues in large-scale sensor networking.
This system integrates a sensor platform, a datastore, and a compute farm.  
The sensor platform is a custom-designed, but inexpensive, sensor platform (called the \ST) paired with general-purpose small-scale compute nodes, including everything from low-power PDAs to powerful touchscreen-based portable computers.  
There are over a dozen sensors packaged on the \ST itself, the board is expandable as well, new sensors can be added to it easily.

% case study: the sensor board development tasks
The case study is focused in building the sensor board and its software driver concurrently.
Because of hard time constraints, and the initial unavailability of the custom board, we need to progress concurrently with all the development tasks, which consist of: (1) specification of the communication protocol; (2) physical board specification; (3) physical board development; (4) development of the board embedded software; (5) communication protocol software driver development; (6) development of software simulators of the board.

% what we have to do, what we outsourced, and the dependencies between tasks
The development of the physical board, along with its embedded software (development tasks 2, 3, 4), is carried out by an external manufacturer, thus it is not directly taken into account here. 
The specification of the communication protocol (development task 1) is a joint effort, while the remaining tasks (5, 6) can be developed in isolation.
The dependencies are straightforward: the specification of the communication protocol (1) is the most stable element, but it still depends on the sensor board (2, 3, 4): big changes in the latter could affect the former.
The driver and the simulators depends directly on the communication protocol.

% communication with external manufacturer
Communication and synchronization with the external manufacturer is frequent but not optimal.
The main synchronization artifact is the protocol specification, that remain stable at high-level, but is changed frequently in the low-level details.
Since domain expertise is very different, a common domain language is hard to find and to agree upon.
Formal methodologies and programming platforms also differ.
The informal specification of the communication protocol is the most reliable source of information.

% how the problem is usually managed
There exist various approaches in literature addressing this kind of development constraints (rapid development in hardware software
co-design), but they are all focused on big-scale systems.
The hardware software formal co-design methodologies usually have several common development artifacts~\cite{Slomka2000,Hoffman2001}: an high-level specification of the system, a translation (refinement) of the high-level specification to low-level ones (hardware and software counterparts), the possibility to generate software code from the low-level specifications, a hardware simulator capable of simulating an hardware component based on hardware specification, the hardware component developed.
When considering the development of the device and of its software driver as separate entities that possibly have to be developed concurrently, the existing approaches are similar to the ones seen in the case of hardware software co-design~\cite{Valderrama1995,Siegmund2002,Ryzhyk2009}: they all focus on a specification that aims to be as complete as possible.

% complete specification is not feasible
A specification as complete as possible is not feasible nor convenient in this case: the protocol specification have to be enhanced incrementally, and its complete specification cannot be provided.
This makes the problem an ideal candidate to test the Mole practices.

% formal methodology constraints
An appropriate formal methodology needs to be chosen; it must be able to support a specification that is built incrementally and accompanied by tools that allow the verification of the system properties at run time.
All the formal methods that require a complete specification to be used effectively must be avoided.
The formal language should also be able to automatically generate tests.

% the chosen formal methodology
Being Java the language chosen for the implementation, the Java Modeling Language (JML)~\cite{advancedJMLpaper}, accompanied by the JML Runtime Assertion Checker suite of tools~\cite{Cheon-Leavens02b}\footnote{\myhref{http://sourceforge.net/projects/jmlspecs/}{Common JML tools}} has been selected.
JML and RAC tools provide a run time checker, a test generator, and do not require a complete specification to be used effectively.

@@

% JML introduction
The Java Modeling Language (JML)~\cite{JML2}\footnote{\myhref{http://www.cs.ucf.edu/~leavens/JML/}{The Java Modeling Language (JML)}} is the de facto standard specification language for describing the behavioural properties of Java programs.  
JML is a rich behavioural interface specification language (BISL) and thus focuses on the modular specification of classes and objects in which inheritance is used solely for behavioural subtyping~\cite{X}. 
JML includes standard specification constructs like assumptions, assertions, axioms, and pre- and postconditions, framing clauses, and invariants. 
It also includes a rich model-based specification layer that includes mathematical models, data refinements~\cite{X}, datagroups~\cite{X}, and many other advanced object-oriented specification features~\cite{advancedJMLpaper}. 
Many tools, ranging from compilers to static checkers to full-functional verification systems support the JML language~\cite{BurdyEtal05-STTT}.

% using JML
JML is sometimes used in a Design-by-Contract style, where a specification is written from scratch, reusing existing APIs that have specifications of their own, and then an implementation is written conforming to that specification~\cite{DBCjml,DBC}.  
The implementation is checked by whatever means appropriate using some subset of the aforementioned tools.  
At other times an existing piece of implemented and tested code is annotated with specifications after-the-fact, a process called Contracting-the-Design.

@@

%\bibliographystyle{latex8} 
%\bibliographystyle{lniguide} 
\bibliography{%
  extra,%
  abbrev,%
  ads,%
  category,%
  complexity,%
  hypertext,%
  icsr,%
  knowledge,%
  languages,%
  linguistics,%
  meta,%
  metrics,%
  misc,%
  modeling,%
  modeltheory,%
  reuse,%
  rewriting,%
  softeng,%
  specification,%
  ssr,%
  technology,%
  theory,%
  web,%
  upcoming,%
  upcoming_conferences,%
  conferences,%
  workshops,%
  verification,%
  escjava,%
  jml,%
  nijmegen}

>>>>

Additionally, in the following when we informally say ``...development
of a system,'' we more precisely mean ``...analysis, design,
development, verification (both testing and formal verification), and
validation of a system.''

@@

This software driver has been developed using our process while
statisfying seemingly contradictory requirements: (1)~the protocol
implementation must be thoroughly verified; (2)~parts of the protocol
must be formally specified; (3)~a board simulator must be designed,
implemented, and verified to match the specifications of, and actual
behavior of, the physical circuit board; (4)~if board components
change during development, the associated formal specifications and
the simulator whose behavior reflects such changes must be updated
accordingly; (5)~strict software delivery deadlines.  Consequently,
these process requirements constitute an environment that benefits
from both an agile and a formal approaches to software development.

@@

\section{A Real and Complex Case Study: \\ Rapid Development in Small-scale Hardware Software Co-design}
\label{subsec:a_real_and_complex_case_study}

@@

\subsection{The UCD CASL SenseTile System}
\label{sec:ucd-casl-sensetile}

% SenseTile introduction
The \textbf{UCD CASL SenseTile System} is a large-scale,
general-purpose sensor system developed at the University College
Dublin in Dublin, Ireland.  Our facility provides a scalable and
extensible infrastructure for performing in-depth investigation into
both the specific and general issues in large-scale sensor networking.
This system integrates a sensor platform, a datastore, and a compute
farm.  The sensor platform is a custom-designed but inexpensive sensor
platform (called the \ST) paired with general-purpose small-scale
compute nodes, including everything from low-power PDAs to powerful
touchscreen-based portable computers.  Besides containing over a dozen
sensors packaged on the \ST itself, the board is expandable as well,
as new sensors are added to it easily.

% case study: the sensor board development tasks
The case study is focused in building the sensor board and its
software driver concurrently.  Because of hard time constraints and
the initial unavailability of the custom board, we must progress
concurrently with all the development tasks, including:
(1)~specification of the communication protocol; (2)~specification of
the physical sensor board; (3)~development and fabrication of the
physical board; (4)~development of the embedded software for the
board; (5)~development of the communication protocol software driver
the the board; and (6)~development of a software simulator of the
board.

% what we have to do, what we outsourced, and the dependencies between tasks
The development of the physical board, along with its embedded
software (development tasks~2, 3, 4), is carried out by an third party
under our guidance, thus it is not directly taken into account here.
The specification of the communication protocol (development task~1)
is a joint effort between ourselves and the third party, while the
remaining tasks~(5, 6) are performed in isolation.  Dependencies are
straightforward: the specification of the communication protocol~(1)
is the most stable element, but it still depends on the sensor
board~(2, 3, 4): large changes in the latter affect the former.
Additionally, the driver and the simulator depends directly on the
communication protocol.

% communication with external manufacturer
Communication and synchronization with the external manufacturer is
frequent but not optimal.  The main synchronization artifact is the
protocol specification, which remains stable at high-level, but is
changed frequently in the low-level details.  Since domain expertise
is very different between the two teams, a common domain language is
difficult to find and to agree upon.  Formal methodologies and
programming platforms also differ.  Consequently, more for social and
psychological reasons than technical ones, the informal specification
of the communication protocol is the most reliable source of
information.

\subsubsection{Related Methodologies}
\label{sec:relat-meth}

% how the problem is usually managed
There exist various approaches in literature addressing this kind of
development constraints (rapid development in hardware software
co-design), but all are focused on large-scale systems.

The hardware-software formal co-design methodologies usually have
several common development artifacts~\cite{Slomka2000, Hoffman2001}
including a high-level specification of the system, a translation
(refinement) of the high-level specification to low-level ones
(hardware and software counterparts), the possibility to generate
software code from the low-level specifications, a hardware simulator
capable of simulating an hardware component based on hardware
specification, the hardware component developed.

When considering the development of the device and of its software
driver as separate entities that possibly have to be developed
concurrently, the existing approaches are similar to the ones seen in
the case of hardware-software co-design~\cite{Valderrama1995,
  Siegmund2002, Ryzhyk2009}, all of which focus on a specification
that aims to be as complete.

% complete specification is not feasible
A complete specification is not feasible nor convenient in our case.
The protocol specification must be enhanced incrementally and its
complete specification cannot be provided.  This makes the problem an
ideal candidate to test the our ``mole'' of practices.

% formal methodology constraints
To perform this case study, an appropriate formal methodology must be
chosen.  It must be able to support a specification that is built
incrementally.  Consequently, formal methods that demand a complete
specification to be used effectively must be avoided.  Additionally,
to best synthesize with agile methods, the formal language must be
able to automatically generate tests.  Finally, the formal method must
be complemented by tools that support the verification of system
properties both at design-time (via static analysis, for full
functional verification of critical subsystems) and at run time (for
dynamic validation for integration with TDD).

% the chosen formal methodology
Driven by these requirements and our expertise, the Java programming
language and platform were chosen for the implementation, the Java
Modeling Language (JML)~\cite{LeavensBakerRuby-Prelim} and Business
Object Notation (BON)~\cite{WaldenNerson95} are used to write formal
specifications, and the BONc~\cite{BONc}, Beetlz~\cite{Beetlz},
JML2~\cite{JML2,JML2common}, and Extended Static Checker for Java
(ESC/Java2)~\cite{ESCJava2,KiniryCok04SmallCite} tool suites are used
as supporting tools~\cite{BurdyEtal05-STTT}.  The JML2 tool suite
includes runtime assertion checker~\cite{Cheon-Leavens02b} and a unit
test generator~\cite{Cheon2002}.  The BONc tool supports validity
checking of high-level requirements and architecture specifications,
Beetlz supports reasoning about refinement between BON and JML
specifications, and ESC/Java2 supports automatic, modular, static
reasoning about the correctness of an implementation against its
formal specification.  All four tool suites do not require a complete
specification to be used effectively.

@@

% JML introduction
The Java Modeling Language (JML) is the de facto standard
specification language for describing the behavioural properties of
Java programs.  JML is a rich behavioural interface specification
language (BISL) and thus focuses on the modular specification of
classes and objects in which inheritance is used solely for
behavioural subtyping~\cite{LeavensBakerRuby99}.  JML includes
standard specification constructs like assumptions, assertions,
axioms, and pre- and postconditions, framing clauses, and invariants.
It also includes a rich model-based specification layer that includes
mathematical models, data refinements~\cite{}, datagroups~\cite{}, and
many other advanced object-oriented specification
features~\cite{chalin06:_beyon_asser}.  Many tools, ranging from
compilers to static checkers to full-functional verification systems
support the JML language~\cite{BurdyEtal05-STTT}.

% using JML
JML is sometimes used in a Design-by-Contract style, where a
specification is written from scratch, reusing existing APIs that have
specifications of their own, and then an implementation is written
conforming to that specification~\cite{Meyer92b}.  The
implementation is checked by whatever means appropriate using some
subset of the aforementioned tools.  At other times an existing piece
of implemented and tested code is annotated with specifications
after-the-fact, a process called Contracting-the-Design.

@@

%\bibliographystyle{latex8} 
%\bibliographystyle{lniguide} 
\bibliography{extra,abbrev,ads,category,complexity,hypertext,icsr,knowledge,languages,linguistics,meta,metrics,misc,modeling,modeltheory,reuse,rewriting,softeng,specification,ssr,technology,theory,web,upcoming,upcoming_conferences,conferences,workshops,verification,escjava,jml,nijmegen}
